---
title: "02 - Informal Analysis"
author: "Randi L. Garcia"
output: html_document
---

# Loading data

```{r, warning=FALSE, message=FALSE}
library(tidyverse)
library(mosaic)
library(readr)

animals <- read_csv("https://randilgarcia.github.io/sds290fall21/animals.csv")

animals <- animals %>%
  mutate(student = seq.int(1, nrow(animals))) %>%
  select(student, order, animal, cute_calm, scary_calm, cute_happy, scary_happy, 
         cute_nerv, scary_nerv, cute_scared, scary_scared) %>%
  mutate_at(vars(cute_calm:scary_scared), as.numeric) %>%
  slice(3:length(animals)) %>% 
  filter(!is.na(cute_calm), !is.na(scary_calm)) %>%
  gather(var, value, cute_calm:scary_scared) %>%
  separate(var, c("cue", "emotion"), sep = "_") %>%
  spread(emotion, value) %>%
  unite(cell, cue, animal, remove = FALSE) %>%
  select(student, animal, cue, cell, calm) 
```

# Visualizing the data

We can use a side-by-side boxplot.

```{r}
#looking at variability by condition
ggplot(animals, aes(x = animal, fill = cue, y = calm)) +
  geom_boxplot()
```

Or parallel dot graph

```{r}
ggplot(animals, aes(x = animal, y = calm, color = cue)) +
  geom_jitter(height = 0.00, width = 0.05, alpha = .7)
```

# Assembly Line Metaphor

## Assembly Line Instructions

To make the set of assembly instructions, first we calculate the effects of all of our structural factors under the Fisher assumptions. We use the data to calculate all of these effects.

The chunks of code below calculate all of the pieces we'll need in our assembly line. Each chunk is a station. The ANOVA conditions assumptions are built into the way we are calculating our assembly line instructions and when simulating the data below. 

```{r}
animals_sim <- animals %>%
  mutate(benchmark = mean(calm)) #benckmark is the grand average or grand mean
```

```{r}
animals_sim <- animals_sim %>%
  group_by(animal) %>%
  mutate(animal_mean = mean(calm),
         aminal_effect = animal_mean - benchmark) #for each level of animal factor
```

```{r}
animals_sim <- animals_sim %>%
  group_by(cue) %>%
  mutate(cue_mean = mean(calm),
         cue_effect = cue_mean - benchmark) #for each level of cue factor
```

```{r}
animals_sim <- animals_sim %>%
  group_by(cell) %>%
  mutate(cell_mean = mean(calm),
         interaction_effect = cell_mean - (animal_mean + cue_mean - benchmark)) #for each cell of interaction
```

```{r}
animals_sim <- animals_sim %>%
  group_by(student) %>% 
  mutate(student_mean = mean(calm),
         student_effect = student_mean - animal_mean) #for each level of student factor, nuisance factor
```

```{r}
animals_sim <- animals_sim %>%
  ungroup() %>%
  mutate(residuals = calm - (benchmark 
                             + aminal_effect 
                             + cue_effect 
                             + interaction_effect 
                             + student_effect)) %>% #universal factor
  select(-animal_mean, -cue_mean, 
         -cell_mean, -student_mean) #removing the stuff we don't need
  
```

Lastly, we create a blank calm variable such that we can simulate it next.  

```{r, simulation}
animals_sim <- animals_sim %>%
  mutate(calm_sim = NA) #initializing simulated variable
```

## Simulating Data

Next, we can simulate new calm data using our effect size calculations, with some assumptions in mind. 

```{r}
animals_sim <- animals_sim %>%
  mutate(calm_sim = benchmark 
         + aminal_effect 
         + cue_effect 
         + interaction_effect 
         + student_effect
         + rnorm(82, 0, 0.79)) #SIMULATED residuals factor
```

## Comparing Simulated Data to Real Data

Let's visualize the data

### Parallel dot graph

```{r}
ggplot(animals_sim, aes(x = animal, y = calm_sim, color = cue)) +
  geom_jitter(height = 0, width = 0.03, alpha = .7)
```

```{r}
ggplot(animals, aes(x = animal, y = calm, color = cue)) +
  geom_jitter(height = 0.10, width = 0.05, alpha = .7)
```

### Side-by-side Boxplots

```{r}
ggplot(animals_sim, aes(x = animal, y = calm_sim, color = cue)) +
  geom_boxplot()
```

```{r}
ggplot(animals_sim, aes(x = animal, y = calm, color = cue)) +
  geom_boxplot()
```


