---
title: "05 - Formal ANOVA"
author: "Randi L. Garcia"
date: "September 14, 2021"
output: html_document
---

# Loading data

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(mosaic)
library(ggplot2)

leaf <- data.frame(diet = c("control", "control", "sucrose", "sucrose", 
                            "glucose", "glucose", "fructose", "fructose"), 
                   days = c(2.3, 1.7, 3.6, 4.0, 3.0, 2.8, 2.1, 2.3))
```

# Visualize the means

```{r}
leaf %>%
  group_by(diet) %>%
  summarise(means = mean(days)) %>%
  mutate(bench = mean(means)) %>%
ggplot(aes(x = diet, y = means)) +
  geom_bar(stat = "identity") +
  geom_hline(aes(yintercept = bench), color = "red") 
```

```{r}
#although only two observations per condition
leaf %>%
  group_by(diet) %>%
  summarise(sd = sd(days))
```


# Calculating effects to go in factor diagram

```{r}
leaf <- leaf %>%
  mutate(benchmark = mean(days)) %>%
  group_by(diet) %>%
  mutate(grp_mean = mean(days),
         diet_effect = grp_mean - benchmark) %>%
  ungroup() %>%
  mutate(fitted = benchmark + diet_effect,
         resid = days - grp_mean,
         resid_alt = days - fitted)
```

# Formal ANOVA

ANOVA for One-Way design

```{r}
mod <- lm(days ~ diet, data = leaf)

anova(mod)
```

# Pairwise tests

You will learn much more about what are called **post hoc tests** next week. But basically, we can do pairwise comparisons, think independent samples t-test between conditions.

For example:

```{r}
leaf %>%
  filter(diet == "control" | diet == "sucrose") %>%
  t.test(days ~ diet, data = .)

leaf %>%
  filter(diet == "control" | diet == "sucrose") %>%
  t.test(days ~ diet, data = ., var.equal = TRUE)
```

Or using the linear model from above

```{r}
summary(mod)
confint(mod)
```

We probably want to adjust for multiple comparisons to control our **experimentwise type I error rate**. 

```{r}
TukeyHSD(mod, "diet")

4.992^2
-4.9923

24.923
```


