---
title: "BF[1] in R"
author: "Randi L. Garcia"
output: html_document
---

# Loading data

```{r, warning=FALSE, message=FALSE}
library(dplyr)
library(mosaic)
library(ggplot2)

parks <- data.frame(passes = c(rep(0, 4), rep(25, 4), 
                            rep(75, 4), rep(200, 4), 
                            rep(500, 4)), 
                    height = c(20.7, 15.9, 17.8, 17.6, 
                            12.9, 13.4, 12.7, 9.0, 
                            11.8, 12.6, 11.4, 12.1, 
                            7.6, 9.5, 9.9, 9.0, 
                            7.8, 9.0, 8.5, 6.7))

parks <- parks %>%
  mutate(passes = as.factor(passes))

```

# Informal Analysis

```{r}
parks %>%
  group_by(passes) %>%
  summarise(mean = mean(height),
            sd = sd(height)) 
```

```{r}
2.022/0.506
```

We might consider a transformation. The ratio of largest to smallest SD is greater than 3. But it only gets worse when I try log, or sqrt. It doesn't look so bad when we make a parallel boxplot, and the sd does not seem related to the mean. 

```{r}
ggplot(parks, aes(x = passes, y = height)) +
  geom_point()
```


# Formal ANOVA

ANOVA for BF[1] design

```{r}
mod <- lm(height ~ passes, data = parks)

anova(mod)
```

Testing residuals

```{r}
parks <- parks %>%
  mutate(residuals = residuals(mod),
         SD = sqrt(2.062))
```

```{r}
ggplot(parks, aes(x = residuals)) +
  geom_histogram(bins = 10) +
  geom_vline(aes(xintercept = SD), color = "blue") +
  geom_vline(aes(xintercept = -SD), color = "blue")
```

By the empirical rule, about 2/3 of the residuals should be within 1 standard deviation. 

# ANOVA as a Linear Model

```{r}
parks <- parks %>%
  mutate(trampled = ifelse(passes == 0, 0, 1))
```

```{r}
parks %>%
  group_by(trampled) %>%
  summarise(mean = mean(height))
```


```{r}
mod2 <- lm(height ~ trampled, data = parks)

summary(mod2)
anova(mod2)
```

```{r}
t.test(height ~ trampled, data = parks, var.equal = TRUE)
```

They all have the same p-value. Further, the t-value (in both t-tests) is the `sqrt()` of the F-value. 

```{r}
sqrt(42.476)
```
