---
title: "Piglets Data BF[2]"
author: "Randi Garcia"
---
  
```{r global-options, include=FALSE}
knitr::opts_chunk$set(eval = FALSE)
library(dplyr)
library(lattice)
library(ggplot2)
library(tidyr)
library(dae)
library(mosaic)
```

#Enter the data

```{r}
piglets <- data.frame(gain = c(1.3, 1.19, 1.08, 
                               1.05, 1.0, 1.04,
                               1.26, 1.21, 1.19,
                               1.52, 1.56, 1.54), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))
```

#Explore the Data

First we get the cell counts. 

```{r}
tally(~ B12|antibiotic, data = piglets)
```

Then we look at it visually. 

```{r}
ggplot(piglets, aes(x = gain)) + geom_histogram(bins = 6)

ggplot(piglets, aes(x = group, y = gain)) + 
  geom_boxplot()
```

As an alternative to the `ggplot()` code above using `group`. 

```{r}
ggplot(piglets, aes(x = interaction(B12, antibiotic), y = gain)) + 
  geom_boxplot()
```

Create a parallel dot plot for piglets data

```{r}
ggplot(piglets, aes(x = B12, y = gain, shape = antibiotic, color = antibiotic)) +
  geom_point(size = 2.5)

ggplot(piglets, aes(x = B12, y = gain, fill = antibiotic)) +
  geom_boxplot()
```

#Check for Equal Variances

```{r}
favstats(~gain, data = piglets)
favstats(gain~group, data = piglets)[,c(1,7:8)]

favstats(gain ~ antibiotic|B12, data = piglets)[,c(1,7:8)]
favstats(gain ~ B12|antibiotic, data = piglets)[,c(1,7:8)]
```

Note variances unequal. Try log transformation. p 217 in text: transformation made to make hand calculations easier.

```{r}
piglets <- piglets %>%
  mutate(lgain = log10(gain),
         new_gain = (gain-1)*100) 

favstats(lgain ~ antibiotic|B12, data = piglets)[,c(1,7:8)]

ggplot(piglets, aes(x = interaction(B12, antibiotic), y = lgain)) + 
  geom_boxplot()
```


```{r}
favstats(new_gain ~ antibiotic|B12, data = piglets)[,c(1,7:8)]

piglets %>%
  group_by(antibiotic, B12) %>%
  summarize(mean = mean(new_gain),
            sd = sd(new_gain))

ggplot(piglets, aes(x = interaction(B12, antibiotic), y = new_gain)) + 
  geom_boxplot()
```

#Formal Analysis

Run two-way ANOVA without the interaction term.

```{r}
mod1 <- lm(new_gain ~ antibiotic + B12, data = piglets)

anova(mod1)
```

##Interactions

Check for interactions: Create an interaction plot.

```{r}
ggplot(piglets, aes(x = B12, y = new_gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_jitter(height = 0, width = 0.03, alpha = .7) +
  geom_smooth(method = "lm", se = 0)
```

Conduct a two-way ANOVA

```{r}
mod2 <- lm(new_gain ~ antibiotic * B12, data = piglets)

anova(mod2)
```

#Residual Analysis 

Check conditions on the residuals

```{r}
par(mfrow=c(1,3))
hist(mod2$residuals)
qqnorm(mod2$residuals)
plot(mod2$fitted.values,mod2$residuals)
```

