---
title: "Confidence Intervals and Effect Sizes"
author: "SDS 290"
date: "September 21, 2021"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, include = FALSE}
library(Stat2Data)
library(tidyverse)
data("Diamonds")
ds <- Diamonds %>%
  group_by(Color) %>%
  summarise(n = n(),
            m = mean(TotalPrice),
            sd = sd(TotalPrice))
```

## Announcements

- Grades are posted for HW2
    - See feedback in comments!
- HW3 posted    
    - 5.54d: “Compare your answers to parts (b) and (c). Discuss the similarities and differences between the two methods.”
    - 5.60 a and b: 
    
```{r}
NCbirths_no7 <- NCbirths %>%
  filter(RaceMom != 7)
```
    
- In-person Office hours (Bass 412)
    - Thurs 1:20-2:35p
    - or by appointment
- Zoom Office hours are full
- Where to get HW help
    - [Spinelli center](https://www.smith.edu/qlc/tutoring.html) tutoring Sun-Thurs 7-9p, Sabin-Reed 301.

## Agenda

- CI and Effect sizes warm-up
- Multiple comparisons
- MP1 topics time
- Introduction to Qualtrics

## Last time

### Effect Sizes and Confidence Intervals

- The p-value for our F-ratio only tells us if effects are **detectable**.
    - Answers the question: **IS there an effect?** 
- Practical significance is different from **statistical** significance.
    - Need to answer the question: **How big is IT**? 
- Which values in ANOVA do we want to construct confidence intervals for (or calculate an effect size for)? 

#### Some options:

  - A specific group mean. 
  - How far away a specific condition is from the grand average (the treatment effect).
  - **How far away a specific condition is from another condition.** 
  - How far away a specific condition (or set of conditions) is(are) from a(nother) set of conditions.
  - Additional effect size option: The overall variance explained by a factor of interest.

## Calculating CI between groups in R

$$(\bar{y_G}-\bar{y_F}) \pm t^*\cdot SD \sqrt{1/n_G+1/n_F}$$

For difference between G (near colorless) and F (colorless)

```{r}
MSE = 0.6771 #from our ANOVA source table
df_E = 344 #from our ANOVA source table
t <- qt(.975, df_E) #for 95% CI

n_g = filter(ds, Color == "G")$n #sample size for G
n_f = filter(ds, Color == "F")$n #sample size for F

mean_g <- log(filter(ds, Color == "G")$m) #mean for G
mean_f <- log(filter(ds, Color == "F")$m) #mean for F

#Confidence interval
UL <- (mean_g-mean_f) + t*sqrt(MSE)*sqrt(1/n_g+1/n_f) #upper limit
LL <- (mean_g-mean_f) - t*sqrt(MSE)*sqrt(1/n_g+1/n_f) #lower limit
```

We are 95% confident that the true mean difference in (log) price between near colorless and colorless Diamonds is between -0.20 and 0.29.   


## Effect Size 

But **how big is IT**?

### Effect Size (D)

$$D_{ij} = \frac{(\bar{y_i}-\bar{y_j})}{SD}$$

- Where again $SD = \sqrt{MSE}$ 

```{r}
D = (mean_g-mean_f)/sqrt(MSE)
D
```

The difference between the price of nearly colorless diamonds and colorless diamonds is only 0.052 times the size of the typical within-group deviations in price. Such a small difference!


## Effect Size 

### R-squared for One-Way ANOVA

But how much does color matter overall?

$R^2$ is the proportion of total variation explained by the factor:

$$R^2 = \frac{SS_{treatment}}{SS_{Total}}$$

OR

$$R^2 = 1-\frac{SS_{error}}{SS_{total}}$$

## R-squared for Diamond Color

```{r}
modDia <- lm(log(TotalPrice) ~ Color, data = Diamonds)
anova(modDia)
```

```{r}
R2 = 29.946/(29.946+232.913)
R2
```

The $R^2$ value is 0.11, so about 11% of the variation in the total price of a diamonds is explained by the diamonds' colors. 

## Try it!

Use the `SandwichAnts` dataset to calculate the CI and effect size for difference between `MultiGrain` and `Rye` in number of `Ants`. Can calculate the $R^2$ for `Bread` and interpret it in a sentence. 

## Multiple Comparisons

## Mini Project 1

Topic brainstorm
