---
title: "Introduction to Blocking"
author: "SDS 290"
date: "October 5, 2021"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(message = FALSE, warning = FALSE)
```

```{r, include = FALSE}
library(Stat2Data)
library(tidyverse)
data("Diamonds")
ds <- Diamonds %>%
  group_by(Color) %>%
  summarise(n = n(),
            m = mean(TotalPrice),
            sd = sd(TotalPrice))
```

## Announcements

- Grades for HW4 posted
- MP1 pre-approval due Friday 11:55p  
- Office hours have changed!
    - Wed 10:50-12:05p (in-person)
    - Thurs office hours **permanently cancelled**!
    - Friday office hours 12:15p-3:00p (**expanded**)
        - Please only sign up once per week
        - Fri appointment slots will open on Thurs morning each week
- Where to get MP1 help
    - Post questions to #mini-project1 channel on Slack!
    - Direct message me on Slack.

## MP1 Pre-Approval

Your completed Qualtrics experiments are due for Pre-Approval by **Fri Oct 8 11:55p**.

  1. Export your survey as a Word doc. Include the survey flow and any skip logic you used. 
  2. Upload the Word doc to Moodle. 

I should be able to see any images you used or links to media. I'll be paying particular attention to your consent form. It's not unusual for me to ask for revisions before you launch your study. 

## Next steps:

I will give you approval or revisions via Slack direct messages. Watch out for those!

Once you have completed any revisions and receive approval from me you can complete the following **checklist** and then share your anonymous survey link. 

  1. Proof read for errors.
  2. Test (preview) your survey multiple times, creating fake data.
  3. [Download the data](https://www.qualtrics.com/support/survey-platform/data-and-analysis-module/data/download-data/export-data-overview/) and make sure it makes sense. 
  4. Zero out the [randomization counts](https://www.qualtrics.com/support/survey-platform/survey-module/survey-flow/standard-elements/randomizer/#EditingCounts) in the survey flow. 
  5. [Delete](https://www.qualtrics.com/support/survey-platform/data-and-analysis-module/data/recorded-responses/#DeletingResponses) all of the fake responses. 
  6. [Upgrade](https://www.qualtrics.com/support/survey-platform/sp-administration/coupon-codes/#UpgradingasanExistingUser) your Qualtrics account with the code. 
  6. Activate your survey.
  7. Get the [anonymous survey link](https://www.qualtrics.com/support/survey-platform/distributions-module/distributions-overview/)! 
    
## Agenda

1. Introduction to Blocking
2. Time for MP1

## Example

Modern zoos try to reproduce natural habitats in their exhibits as much as possible. They try to use appropriate plants, but these plants can be infested with inappropriate insects. Cycads (plants that look vaguely like palms) can be infected with mealybugs, and the zoo wishes to test three treatments: 1) water, 2) horticultural oil, and 3) fungal spores in water. Nine infested cycads are taken to the testing area. Three branches are randomly selected from each tree, and 3 cm by 3 cm patches are marked on each branch. The number of mealybugs on the patch is counted. The three treatments then get randomly assigned to the three branches for each tree. After three days the mealybugs are counted again. The change in number of mealybugs is computed ($before-after$).

[Image of cycads](https://www.huntington.org/sites/default/files/frontiers/images/cycadpassion_1.jpg)

## Example - One-Way Design

Modern zoos try to reproduce natural habitats in their exhibits as much as possible. They try to use appropriate plants, but these plants can be infested with inappropriate insects. Cycads (plants that look vaguely like palms) can be infected with mealybugs, and the zoo wishes to test three treatments: 1) water, 2) horticultural oil, and 3) fungal spores in water. Nine infested cycads are taken to the testing area. The number of mealybugs on each tree is counted. The three treatments then get randomly assigned to the three trees each. After three days the mealybugs are counted again. The change in number of mealybugs is computed ($before-after$).

## Example

Male albino laboratory rats are used routinely in many kinds of experiments. This experiment was designed to determine the requirements for protein in their food. Specifically, the experiment is interested in testing five levels of protein (8.68, 12, 15, 18, and 21% of diet). Baby rats were separated into ten groups of 5 to form groups of approximately the same weight. The 5 rats in each group were randomly assigned to each of the 5 conditions. Body weight and food consumption were measured twice weekly, and the average daily weight gain over 21 days was recorded.

## Example - One-Way Design

Male albino laboratory rats are used routinely in many kinds of experiments. This experiment was designed to determine the requirements for protein in their food. Specifically, the experiment is interested in testing five levels of protein (8.68, 12, 15, 18, and 21% of diet). 50 baby rats were randomly assigned to each of the 5 conditions. Body weight and food consumption were measured twice weekly, and the average daily weight gain over 21 days was recorded.

## Example

This experiment is interested in the blood concentration of a drug after it has been administered. The concentration will start at zero, then go up, and back down as it is metabolized. This curve may differ depending on the form of the drug (a solution, a tablet, or a capsule). We will use three subjects, and each subject will be given the drug three times, once for each method. The area under the time-concentration curve is recorded for each subject after each method of drug delivery.   

## Example - One-Way Design

This experiment is interested in the blood concentration of a drug after it has been administered. The concentration will start at zero, then go up, and back down as it is metabolized. This curve may differ depending on the form of the drug (a solution, a tablet, or a capsule). We will use nine subjects, and randomly assign subjects to one of the three delivery methods. The area under the time-concentration curve is recorded for each subject after being given the drug.


## Design Principal: Blocking

- Blocking is using a factor that is not of research interest
    – But there may be differences across blocks on the response variable
- A "block" is a level of a blocking factor
- We use blocking to improve precision/statistical power of our factor of interest


## Three Ways to Block

1. Sort units into similar groups
    - Albino rats
2. Subdivide larger chunks of material into sets of smaller pieces
    - Mealybug trees
3. Reuse subjects or material in each of several times slots
    - Drug study


## Complete Block Design

- Experimental material are separated into groups (or reused) to create similar units 
- Then each unit within a block is then is assigned one level of the factor of interest
- "Complete Block" means that every block x treatment combination is tested

## Inappropriate Insects

Modern zoos try to reproduce natural habitats in their exhibits as much as possible. They try to use appropriate plants, but these plants can be infested with inappropriate insects. Cycads (plants that look vaguely like palms) can be infected with mealybugs, and the zoo wishes to test three treatments: 1) water, 2) horticultural oil, and 3) fungal spores in water. Nine infested cycads are taken to the testing area. Three branches are randomly selected from each tree, and 3 cm by 3 cm patches are marked on each branch. The number of mealybugs on the patch is counted. The three treatments then get randomly assigned to the three branches for each tree. After three days the mealybugs are counted again. The change in number of mealybugs is computed ($before-after$).

## Inappropriate Insects

```{r, echo=FALSE}
library(tidyr)
library(dplyr)
library(knitr)

tree <- c(rep("tree1",3), rep("tree2",3), rep("tree3", 3),rep("tree4", 3), rep("tree5", 3))
treatment <- rep(c("water", "spores", "oil"), 5)
bugs_change <- c(-9, -4, 4, 18, 29, 29, 10, 4, 14, 9, -2, 14, -6, 11, 7)

mealybugs <- data.frame(tree, treatment, bugs_change)

mealybugs %>%
  spread(tree, bugs_change) %>%
  kable()
```

Draw the factor diagram. 

## Formal ANOVA for Complete Block Design

$${y}_{ijk}={\mu}+{\tau}_{i}+{\beta}_{j}+{e}_{ijk}$$

| Source | SS | df | MS | F |
|------:|:-----:|:-----:|:------:|:------:|
| Treatment | $\sum_{i=1}^{a}b(\bar{y}_{i.}-\bar{y}_{..})^{2}$ | $a-1$ | $\frac{{SS}_{T}}{{df}_{T}}$ | $\frac{{MS}_{T}}{{MS}_{E}}$ |
| Blocks | $\sum_{j=1}^{b}a(\bar{y}_{.j}-\bar{y}_{..})^{2}$ | $b-1$ | $\frac{{SS}_{B}}{{df}_{B}}$ | $\frac{{MS}_{B}}{{MS}_{E}}$ |
| Error | $\sum_{i=1}^{a}\sum_{j=1}^{b}({y}_{ij}-\bar{y}_{i.}-\bar{y}_{.j}+\bar{y}_{..})^{2}$ | $(a-1)(b-1)$ | $\frac{{SS}_{E}}{{df}_{E}}$ | |

## Data Analysis Structure

```{r}
mealybugs
```

## Informal Analysis Structure

```{r}
library(tidyr)

mealybugs %>%
  spread(treatment, bugs_change)
```

## Formal ANOVA

```{r}
mod <- lm(bugs_change ~ treatment + tree, data = mealybugs)

anova(mod)
```

Write a sentence interpreting the results of the F-test for treatment and a sentence interpreting the F-test for tree in the context of the problem.

## Formal ANOVA

```{r}
mod <- lm(bugs_change ~ treatment + tree, data = mealybugs)

anova(mod)
```

There are no statistically significant differences in the reduction in mealy bugs between the three treatment conditions, $F(2, 8) = 3.00$, $p = .107$. There are significant differences in the reduction in mealy bugs across trees, however, $F(4, 8) = 9.04$, $p = .005$. That is, some trees improved more than other trees. 

## MP1 time

Raise your hand if you have any questions! 

<!-- ## Scatterplots -->

<!-- ```{r} -->
<!-- library(ggplot2) -->

<!-- mealybugs %>% -->
<!--   spread(treatment, bugs_change) %>% -->
<!--   ggplot(aes(x = spores, y = oil)) + -->
<!--   geom_point() -->
<!-- ``` -->

<!-- ## Scatterplots -->

<!-- ```{r} -->
<!-- library(ggplot2) -->

<!-- mealybugs %>% -->
<!--   spread(treatment, bugs_change) %>% -->
<!--   ggplot(aes(x = spores, y = water)) + -->
<!--   geom_point() -->
<!-- ``` -->

<!-- ## Scatterplots -->

<!-- ```{r} -->
<!-- library(ggplot2) -->

<!-- mealybugs %>% -->
<!--   spread(treatment, bugs_change) %>% -->
<!--   ggplot(aes(x = oil, y = water)) + -->
<!--   geom_point() -->
<!-- ``` -->

<!-- ## Tuplips -->

<!-- A plant breeder wishes to study the effects of soil drainage and variety of tulip bulbs on flower production. Twelve 3m by 10m experimental sites are available in the test garden--each is a .5m deep trench. You can manipulate soil drainage by changing the ratio of sand to clay for the soil you put in a trench. After talking to your collaborator, you decided that four different levels of soil drainage would suffice. You'll be testing 15 different types of tulips, and measuring flower production in the spring.  -->

<!-- ## Bioequivalence of drug delivery -->

<!-- This experiment is interested in the blood concentration of a drug after it has been administered. The concentration will start at zero, then go up, and back down as it is metabolized. This curve may differ depending on the form of the drug (a solution, a tablet, or a capsule). We will use three subjects, and each subject will be given the drug three times, once for each method. The area under the time-concentration curve is recorded for each subject after each method of drug delivery.   -->
