---
title: "Fitting Two-Way Factorial Designs"
author: "SDS 290"
date: "November 9, 2021"
output: 
  ioslides_presentation:
    widescreen: true
    smaller: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(Stat2Data)
```


## Announcements

- HW 6 grades posted; MP1 grades are (mostly) posted
- HW 7 posted and due Friday Nov 12 11:55p
    - Find a research article that uses ANOVA
- In-person Office hours (Bass 412)
    - Wed (tomorrow) 10:50-12:05p 
- Zoom Office hours
    - Friday 12:15-3p
    - Slots open on Thurs morning
- Where to get HW help
    - [Spinelli center](https://www.smith.edu/qlc/tutoring.html) tutoring Sun-Thurs 7-9p, Sabin-Reed 301.
    - Post questions to #homework7-questions channel on Slack!
    - Direct message me on Slack.
- Applications open for DSC-WAV (more info [here](https://dsc-wav.github.io/www/students.html#Eligibility))
    - Due tomorrow
    
## SDS Biostatistics Job Candidates

1. Tea with candidates 3:30-4:30 p.m. Fri and 4:00-4:45 p.m. Mon Wed ([sign-up here!](https://docs.google.com/spreadsheets/d/1uuFRI2TPpYdSWXImIFO9JaXkXgHTihuBtqXOr3NmbjQ/edit#gid=0))
2. Research talks with grab-and-go lunches (12:25-1:20 p.m., Ford 240 or McConnell 103)
3. Teaching demonstrations (5:00-6:00 p.m., Sabin-Reed 301)

**Friday November 12, 2021**: Kaitlyn Cook "Statistical Methods for Cluster-Randomized Infectious Disease Prevention Trial". Research talk: **Ford 240**. 

**Monday November 15, 2021**: Anny-Claude Joseph "The Timing of Geographic Power". Research talk: **Ford 240**. 

**Wednesday November 17, 2021**: Jacob Fiksel "Generalized Bayes Quantification Learning with Applications to Mortality Surveillance". Research talk: **McConnell 103**. 
    
## Agenda

1. MP2 Assignment 
2. Two-Way Factorial Design (continued)
    - More interactions!
    - Fitting the ANOVA model
   
## MP2 Assignment 

- MP2 instructions
- MP2 [group selection survey](https://docs.google.com/forms/d/e/1FAIpQLSdAnm-YO99CPLIkkpaciS9_e0V04bCRFCVsVg0mdHAIOvSJtQ/viewform?usp=sf_link)

## Interaction Graphs

For each of the following interaction graphs, answer the following questions with **YES** or **NO**.

1. Is there a main effect of B12?
2. Is there a main effect of antibiotics?
3. Is there an interaction between B12 and antibiotics?

## Practice 1

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(7, 7, 7, 
                               4, 4, 4,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

## Practice 1

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(7, 7, 7, 
                               4, 4, 4,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

1. NO
2. NO
3. YES

## Practice 2

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               5, 5, 5,
                               6, 6, 6,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

## Practice 2

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               5, 5, 5,
                               6, 6, 6,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

1. YES
2. YES
3. NO

## Practice 3

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               7, 7, 7,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```

## Practice 3

```{r, echo = FALSE, warning=FALSE, message=FALSE}
library(dplyr)
library(ggplot2)
piglets <- data.frame(gain = c(4, 4, 4, 
                               7, 7, 7,
                               4, 4, 4,
                               7, 7, 7), 
                      group = as.factor(c(rep(1, 3), rep(2, 3), rep(3, 3), rep(4, 3))), 
                      antibiotic = c(rep("0mg", 3), rep("40mg", 3), rep("0mg", 3), rep("40mg", 3)),
                      B12 = c(rep("0mg", 6), rep("5mg", 6)))

ggplot(piglets, aes(x = B12, y = gain, 
                    group = antibiotic, 
                    linetype = antibiotic, 
                    shape = antibiotic,
                    color = antibiotic)) +
  geom_point() +
  geom_smooth(method = "lm", se = 0) +
  ylim(2, 9)
```


1. NO
2. YES
3. NO

## Two-Way Factorial Design Formulas

$${y}_{ijk}={\mu}+{\alpha}_{i}+{\beta}_{j}+{\alpha\beta}_{ij}+{e}_{ijk}$$

- Where $i$, from 1 to $a$, is the level of the first factor,
- $j$, from 1 to $b$, is the level of the second factor, 
- and $k$, from 1 to $n$, is the observation in each cell. 

## Sum of Squares (SS)

$${SS}_{A} = \sum_{i=1}^{a}bn(\bar{y}_{i..}-\bar{y}_{…})^{2}$$

$${SS}_{B} = \sum_{j=1}^{b}an(\bar{y}_{.j.}-\bar{y}_{…})^{2}$$

$${SS}_{AB} = n\sum_{i=1}^{a}\sum_{j=1}^{b}(\bar{y}_{ij.}-\bar{y}_{i..}-\bar{y}_{.j.}+\bar{y}_{…})^{2}$$

$${SS}_{E} = \sum_{i=1}^{a}\sum_{j=1}^{b}\sum_{k=1}^{n}({y}_{ijk}-\bar{y}_{ij.})^{2}$$


## Degrees of Freedom (df)

$${df}_{A}=a-1$$

$${df}_{B}=b-1$$

$${df}_{AB}=(a-1)(b-1)$$

$${df}_{E}=ab(n-1)$$

## Mean Squares (MS) 

$${MS}_{A}=\frac{{SS}_{A}}{{df}_{A}}$$

$${MS}_{B}=\frac{{SS}_{B}}{{df}_{B}}$$

$${MS}_{AB}=\frac{{SS}_{AB}}{{df}_{AB}}$$

$${MS}_{E}=\frac{{SS}_{E}}{{df}_{E}}$$


## Three Research Questions, Three F-ratios

1. Is there a significant main effect of factor A?
2. Is there a significant main effect of factor B?
3. Is there an significant interaction between factor A and factor B?

## F-ratios and the F-distribution

The ultimate statistics we want to calculate is Variability in treatment effects/Variability in residuals. The F-ratio.

$$F = \frac{{MS}_{A}}{{MS}_{E}}$$

$$F = \frac{{MS}_{B}}{{MS}_{E}}$$

$$F = \frac{{MS}_{AB}}{{MS}_{E}}$$


## ANOVA Source Table for Two-Way Factorial

$${y}_{ijk}={\mu}+{\alpha}_{i}+{\beta}_{j}+{\alpha\beta}_{ij}+{e}_{ijk}$$

| Source | SS | df | MS | F |
|------:|:-----:|:-----:|:------:|:------:|
| Treatment A | $\sum_{i=1}^{a}bn(\bar{y}_{i..}-\bar{y}_{…})^{2}$ | $a-1$ | $\frac{{SS}_{A}}{{df}_{A}}$ | $\frac{{MS}_{A}}{{MS}_{E}}$ |
| Treatment B | $\sum_{j=1}^{b}an(\bar{y}_{.j.}-\bar{y}_{…})^{2}$ | $b-1$ | $\frac{{SS}_{B}}{{df}_{B}}$ | $\frac{{MS}_{B}}{{MS}_{E}}$ |
| Interaction AB | $n\sum_{i=1}^{a}\sum_{j=1}^{b}(\bar{y}_{ij.}-\bar{y}_{i..}-\bar{y}_{.j.}+\bar{y}_{…})^{2}$ | $(a-1)(b-1)$ | $\frac{{SS}_{AB}}{{df}_{AB}}$ | $\frac{{MS}_{AB}}{{MS}_{E}}$ |
| Error | $\sum_{i=1}^{a}\sum_{j=1}^{b}\sum_{k=1}^{n}({y}_{ijk}-\bar{y}_{ij.})^{2}$ | $ab(n-1)$ | $\frac{{SS}_{E}}{{df}_{E}}$ | |

## Two-Way Factorial in R

[See Two-Way Factorial code](https://randilgarcia.github.io/sds290fall21/lectures/09_two-way_fac_inR.Rmd)

<!-- ## Teaching Methods -->

<!-- A professor wanted to compare three different teaching methods to determine how students  -->
<!-- would perceive the course: 1) instructionist, 2) inquiry-based, and 3) team-based. She randomly assigned the same class (same topic different students) from 6 different semesters to treatments. At the end of the semester students were asked to rate the course on a 5-point scale, and the average class rating was calculated.  -->

<!-- ## Anxiety and Memory -->

<!-- A psychologist wants to study the effect of anxiety on 4 different types of memory. Twelve participants are assigned to one of two anxiety conditions: 1) low anxiety group is told that they will be awarded \$5 for participation and \$10 if they remember sufficiently accurately, and 2) high anxiety group is told they will be awarded \$5 for participation and \$100 if they remember sufficiently accurately. All subjects perform four memory trials in random order, testing 4 different types of memory. The number of errors on each trial is recorded.      -->

