---
title: "R Study Guide"
author: "Prof. Garcia"
date: "Spring 2017"
output: pdf_document
---

1. Given numerical objects named `x` and `y`, calculate this quantity (which is of no special interest): $\sqrt{x^2/y}$   
```{r eval = FALSE}
sqrt(x^2/y)
```

2. Load the `mosaic` and `mosaicData` packages, read in a data set named “CPS85”, and assign it to an object called people.
```{r message = FALSE}
require(mosaic)
data(CPS85)
people <- CPS85
```

3. Display the first few rows of the people data frame. 
```{r eval = FALSE}
head(people)
```

4. Display a quick glance at the data frame.
```{r eval = FALSE}
glimpse(people)
```

5. Calculate (not count by hand!) the number of cases in the data frame.
```{r eval = FALSE}
nrow(people)
```

6. Calculate the mean wage of all the people.
```{r eval = FALSE}
mean(~wage, data = people) 
```

7. Calculate the standard deviation of wage for all cases.
```{r eval = FALSE}
sd(~wage, data = people) 
```

8. Compute the five number summary of all wages with a single command.
```{r eval = FALSE}
favstats(~wage, data = people)
```

9. Calculate the mean wage separately for married and unmarried people.
```{r eval = FALSE}
mean(~wage | married, data = people)
```

10. Create a new variable, `fraction`, in the data frame that holds the ratio of the person’s “experience”
to their age.
```{r eval = FALSE}
people <- people %>%
  mutate(fraction = exper/age)
```

11. Make a box-and-whisker plot of all the people’s wages.
```{r}
qplot(x = married, y = wage, geom = "boxplot", data = people)
```

12. Make a scatterplot of people’s wages as a function of age, separated by marital status:
```{r}
qplot(x = age, y = wage, color = married, data = people)
```

13. Calculate (not count by hand!) the number of people by marital status.
```{r eval = FALSE}
tally(~married, data = people)
```

14. Calculate (not count by hand!) the number of people by marital status and sex simultaneously.
```{r eval = FALSE}
tally(~married | sex, data = people)
```
